Quantitative Benchmarking for Automatically Generated Conjectures

### 2017-10-02 TEBenchmarkPaper/article.tex
Lots of fleshing out

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{hhline}
\usepackage{mathtools}
\usepackage{multirow}
% etc.

@@

\begin{itemize}
\item The quality of each conjecture, which is the ``interestingness'' discussed
  above.
\item The \emph{set} of conjectures produced, to assess things like novelty, as
  well as consistency of the results.
\item Performance of the system, to find the balance struck between output
  quality and time taken.
\end{itemize}

@@

If we limit ourselves to exploring higher-order theories with inductive types
(a domain closely matching our interest in analysing software), we do find a
direct comparison of three systems~\cite{claessen2013automating}: QuickSpec (the
conjecture-generating component of HipSpec), IsaCoSy and IsaScheme. This is a
\emph{precision/recall} analysis, providing the same theories to each system and
comparing the set of generated conjectures against a \emph{ground truth}.

% TODO: precision/recall is evaluating the set of conjectures. It assumes that
% interestingness == membership in the ground truth, and ignores efficiency
% completely

A common choice of ground truth is the standard library of a theorem prover,
such as Isabelle/HOL; the library authors have gone to the effort of stating,
proving and including these theorems in every copy of their software, which is a
good indication that they are useful or important.

@@

We propose a larger, more ambitious method for evaluating ATE systems, which we
call the Theory Exploration Benchmark (TEB). We take a similar approach to prior
evaluations, but our ground truth comes from the extensive test suites which are
already used to benchmark automated \emph{theorem provers}; which we repurpose
for use in the theory \emph{exploration} setting.

@@

\begin{itemize}
\item Each problem provides separate definitions and a theorem statement; hence
  it's not difficult to tease them apart for our purposes.
\item Higher-order functions and inductively defined types are provided, which
  corresponds to our desired domain.
\item All together, TIP provides 219 distinct function definitions and 343
  theorem statements, which is enough to fully exercise current ATE systems.
\item Benchmark problems include examples from the theorem proving literature
  as well as common program verification tasks, ensuring the resulting corpus is
  relevant to researchers and practitioners.
\item TIP's maintainers provide a set of tools, including translators from TIP's
  native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
  (which are used by ATE systems).
\end{itemize}

From TIP, we automatically derive our Theory Exploration Benchmark (TEB), and
provide tools to benchmark ATE systems and compare their results. Our
methodology is as follows:

@@

@@

>>>>

% Derived from the template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
%
% \usepackage{mathptmx}      % use Times fonts if available on your TeX system
%
% insert here the call for the packages your document requires
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{array}
\usepackage{hhline}
\usepackage{mathtools}
\usepackage{multirow}
\usepackage{tikz}
\usetikzlibrary{arrows, calc, decorations.pathreplacing, positioning,
  shapes.geometric}
% etc.

@@

\begin{enumerate}
\item The quality of each conjecture, which is the ``interestingness'' discussed
  above.
\item The quality of the \emph{set} of conjectures produced, used for relative
  interestingness criteria like novelty, as well as to assess consistency.
\item Performance of the system, to find the balance struck between output
  quality and time taken.
\end{enumerate}

@@

If we limit ourselves to exploring higher-order theories with inductive types
(a domain closely matching our interest in analysing software), we do find a
direct comparison of three systems~\cite{claessen2013automating}: HipSpec,
IsaCoSy and IsaScheme. QuickSpec, shown in table \ref{table:colton}, is the
conjecture-generating component of HipSpec (which sends those conjectures into
automated theorem provers).

This comparison uses \emph{precision/recall analysis}: a particular set of
definitions is chosen, such as those in figure \ref{figure:list_theory}, and a
\emph{ground truth} is chosen as the 'ideal' set of conjectures which we would
like an ATE system to produce from these definitions.

Each system is run, and their output compared to the ground truth. To score
100\% on precision and recall, an ATE system must output all of the conjectures
which appear in the ground truth, and nothing else:

\begin{itemize}
\item \emph{Precision} is the proportion of a system's output which appears in
  the ground truth. This penalises systems which output large numbers of
  conjectures in the hope that some turn out to be ``good''.
\item \emph{Recall} is the proportion of the ground truth which appears in the
  system's output. This penalises systems which overly restrict their output to
  avoid generating ``bad'' conjectures.
\end{itemize}

Precision and recall fulfil our second requirement, by directly measuring the
quality of the \emph{set} of conjectures. Their ``interestingness'' criterion is
indirect: asking only whether or not a conjecture appears in the ground truth.
A satisfactory solution to our first requirement hence requires careful choice
of the ground truth set.

This particular analysis takes its ground truths from the standard library of
the Isabelle theorem prover, with one involving the natural numbers, and the
other involving the list functions from figure \ref{figure:list_theory}. The library authors have gone to the effort of stating,
proving and including these theorems in every copy of their software, which is a
good indication that they are useful or important.

@@

Our main contribution is a benchmarking methodology, shown in figure
\ref{figure:flow_chart}, which generates both a large definition/ground-truth
corpus (an order of magnitude larger than previous work) and a scalable,
statistical approach to evaluating ATE systems using this corpus. We follow a
precision/recall approach similar to the work described above, with the main
difference being the source of definitions and ground truths: we take existing
problem sets designed for automated \emph{theorem proving}, and adapt their
content for use in the \emph{theory exploration} setting.

\subsection{Preparation}

Automated theorem proving is an active area of research, with large problem sets
and regular competitions to prove as much as possible, as fast as possible.
These problem sets are an opportunity for theory exploration, as their
definitions and theorems can be used as a corpus, in the same way that Isabelle
libraries have been used in the past.

Some problem sets are more amenable for this purpose than others. The most
suitable are those which meet the following criteria:

\begin{itemize}
\item For each problem, there should be a clear separation between the theorem
  to be proved and the definitions involved, such that the two can be easily and
  meaningfully separated. This rules out problem sets like SMTLIB, where many
  problems \emph{declare} a function, but leave its behaviour \emph{implicit} in
  the logical structure of the theorem statement.
\item Definitions should be \emph{strongly typed}, such that they can be
  translated into the native language of each ATE system (in this work, those
  languages are Haskell and Isabelle).
\item The problem set should be relevant to the desired domain. In our case, we
  desire higher-order functions and inductively defined types, which rules out
  first-order languages/logics.
\item Since it will act as our ground truth, the problem set should ideally
  contain \emph{every} ``interesting'' conjecture involving its included
  definitions. Realistically, we should aim for each definition to appear in
  \emph{many} theorem statements; rather than each problem having unique
  definitions.
\item The problem set should be as large as possible, for robustness of the
  resulting statistics.
\end{itemize}

Once a theorem proving problem set has been chosen, we must separate the
definitions referenced by the theorems from the theorem statements themselves.
The former will be used as input to the ATE systems under evaluation, whilst the
latter form a ground truth corpus against which to compare the ATE output.

It is important to ensure that there are no duplicate definitions: theory
exploration does not depend on the names of functions, so neither should our
analysis. For example, consider a problem set which includes a commutativity
theorem for a \texttt{plus} function, and an associativity theorem for an
\texttt{add} function, where the definitions of \texttt{plus} and \texttt{add}
are $\alpha$-equivalent. We would expect an ATE system to conjecture
commutativity and associativity for \emph{both} functions, or for \emph{neither}
function, since they're equivalent. Yet a na\"ive precision/recall analysis
would treat commutativity of \texttt{add} and associativity of \texttt{plus} as
\emph{uninteresting}, since they don't appear in the ground truth.

For this reason, duplicates should be removed (e.g. leaving those which appear
first alphabetically), and any references in the theorem statements updated to
use the remaining definition. In the above example, the \texttt{plus} function
would be removed, and the commutativity theorem updated to reference
\texttt{add} instead.

\subsection{Sampling}

We could, in theory, send these de-duplicated definitions straight into an ATE
system, and use the updated theorem statements as the ground truth for analysis.
However, this would cause two problems:

\begin{itemize}
\item The result would be a single data point, which makes it difficult to
  infer performance \emph{in general}.
\item It is impractical to run any existing ATE system on ``large'' inputs,
  containing more than around 10 definitions.
\end{itemize}

To solve both of these problems we instead \emph{sample} a subset of
definitions. Given a sample size, we choose a subset of that many definitions,
and use that as our ATE input. We generate a corresponding ground truth by
selecting those theorems from the corpus which only contain references to
definitions that appear in our sample.

Unfortunately, uniform sampling of definitions gives rise to a lottery, where
for a large corpus and a fixed sample size, it becomes highly unlikely that a
sample will contain the particular definitions required by a theorem. This would
cause most samples to have an empty ground truth, and hence $0$ precision and
undefined recall \emph{independent} of what the ATE system produces. This is
clearly undesirable as an evaluation method.

Instead, we only choose samples which admit at least one of the corpus theorems.
We could do this using rejection sampling, but it is more efficient to sample
the \emph{theorem statements}, weighted in proportion to the number of
definitions they reference. Once a theorem is chosen, we use the referenced
definitions as our sample, padding uniformly up to the required size (we ignore
theorems which reference more definitions than our sample size).

We form the ground truth as above: selecting all applicable theorems from the
corpus, which is now guaranteed to be non-empty.

\subsection{Evaluation}

Given a sample of definitions and a corresponding ground truth, the actual
execution of the ATE system proceeds as in prior work. We must translate the
definitions into the required input format, which depends on the particular
tool being studied, then we time the execution, with a cutoff after e.g. 1 hour.

In our experiments we have found that memory usage also plays an important role,
but rather than introducing an extra dimension we can account for this by simply
allowing the system

\begin{figure}
  \tikzstyle{startstop} = [rectangle, rounded corners, minimum width=3cm,
                           minimum height=1cm,text centered, draw=black]

  \tikzstyle{io} = [trapezium, trapezium left angle=70,
                    trapezium right angle=110, minimum width=1cm,
                    minimum height=1cm, text centered, draw=black]

  \tikzstyle{process} = [rectangle, minimum width=3cm, minimum height=1cm,
                         text centered, draw=black]
  \tikzstyle{decision} = [diamond, minimum width=3cm, minimum height=1cm,
                          text centered, draw=black]

  \tikzstyle{arrow} = [thick,->,>=stealth]

  % Avoids too much padding in io shapes
  \tikzset{trapezium stretches=true}

  \centering
  \begin{tikzpicture}[node distance=1cm]

    % Preparation section

    \node (in)  [io]{Theorem Proving Benchmark (TIP)};
    \node (sep) [process,       below=1cm of in         ]{Separate definitions from theorems};
    \node (def) [process, below  left=1.5cm and -1cm of sep]{Remove duplicate definitions};
    \node (ref) [process, below right=1.5cm and -1cm of sep]{Update references};

    \node (thy)  [startstop, below=1cm of def]{Full Theory};
    \node (thm)  [startstop, below=1cm of ref]{Theorem Corpus};

    \draw [arrow] (in)  -- (sep);
    \draw [arrow] (def) -- (ref);
    \draw [arrow] (def) -- (thy);
    \draw [arrow] (ref) -- (thm);

    % Arrows with labels
    \path[->]
        (sep) edge [arrow, sloped, above] node {definitions} (def)
        (sep) edge [arrow, sloped, above] node {theorems}    (ref);

    % Sampling section

    \node (choose) [process, below=of thm   ]{Sample a theorem};
    \node (deps)   [process, below=of choose]{List referenced definitions};

    % Create dummy coordinate below deps, then use its y coordinate for pad
    \coordinate [below=of deps] (padDummy);
    \path let \p{dummy} = (padDummy),
              \p{in}    = (in)
              in coordinate (padPos) at (\x{in}, \y{dummy});
    \node (pad)  [process, at=(padPos)]{Sample definitions to pad list};

    \node (find) [process,   below right=1.5cm and -1cm of pad]{List all applicable theorems};
    \node (sthy) [startstop, below  left=1.5cm and -1cm of pad]{Sampled Theory};

    \node (sthm)  [startstop, below=of find]{Ground Truth};

    % Calculate position of (size) using let, then define as normal
    \path let \p{pad}    = (pad),
              \p{choose} = (choose)
           in coordinate (sizePos) at (\x{pad},\y{choose});
    \node (size) [io, at=(sizePos)] {Sample size};

    % Evaluation section
    \path let \p{pad}  = (pad),
              \p{sthm} = (sthm)
           in coordinate (runPos) at (\x{pad}, \y{sthm});
    \node (run) [process, below=of runPos]{Run ATE system};
    \node (pr)  [process, below=of run]{Analysis};

    \node (prec) [startstop, below=of pr  ]{Precision};
    \node (time) [startstop,  left=of prec]{Time taken};
    \node (rec)  [startstop, right=of prec]{Recall};

    \draw [arrow] (thy)    |- (pad);
    \draw [arrow] (thm)    -- (choose);
    \draw [arrow] (choose) -- (deps);
    \draw [arrow] (deps)   |- (pad);
    \draw [arrow] (size)   -- (choose);
    \draw [arrow] (size)   -- (pad);
    \draw [arrow] (pad)    -- (find);
    \draw [arrow] (pad)    -- (sthy);
    \draw [arrow] (find)   -- (sthm);
    \draw [arrow] (sthy)   |- (run);
    \draw [arrow] (run)    -- (pr);
    \draw [arrow] (sthm)   |- (pr);
    \draw [arrow] (pr)     -- (prec);
    \draw [arrow] (pr)     -- (rec);
    \draw [arrow] (pr)     -- (time);

    % Awkward arrow
    \draw [arrow] (thm) -| ([shift={(7mm,-7mm)}]thm.east) |- (find);

    % Braces

    % Preparation
    \draw
      let \p{thm} = (thm.east),
          \p{in}  = (in.north),
          \p{rec} = (rec.south east)
       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
       (\x{rec}, \y{in}) -- (\x{rec}, \y{thm})
         node [black, midway, right, xshift=0.3cm] {Preparation};

    % Sampling
    \draw
      let \p{thm} = (thm.east),
          \p{gt}  = (sthm.east),
          \p{rec} = (rec.south east)
       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
       (\x{rec}, \y{thm}) -- (\x{rec}, \y{gt})
         node [black, midway, right, xshift=0.3cm] {Sampling};

    % Evaluation
    \draw
      let \p{sthm} = (sthm.east),
          \p{rec}  = (rec.south east)
       in [decorate,decoration={brace, amplitude=10pt}, xshift=0.5cm]
       (\x{rec}, \y{sthm}) -- (\x{rec}, \y{rec})
         node [black, midway, right, xshift=0.3cm] {Evaluation};
  \end{tikzpicture}
  \caption[]{High-level view of our benchmarking methodology, showing
    \begin{tikzpicture}
      \node [rectangle, text centered, draw=black]{processes};
    \end{tikzpicture},
    \begin{tikzpicture}
      \node [rectangle, rounded corners, text centered, draw=black]{data};
    \end{tikzpicture} and
    \begin{tikzpicture}
      \node [trapezium, trapezium left angle=70, trapezium right angle=110,
             text centered, draw=black]{inputs};
  \end{tikzpicture}}
  \label{figure:flow_chart}
\end{figure}

We call the Theory Exploration Benchmark (TEB).

@@

All together, TIP provides 219 distinct function definitions and 343
  theorem statements, which is enough to fully exercise current ATE systems.
Benchmark problems include examples from the theorem proving literature
  as well as common program verification tasks, ensuring the resulting corpus is
  relevant to researchers and practitioners.
TIP's maintainers provide a set of tools, including translators from TIP's
  native format (based on SMT-Lib~\cite{BarFT-SMTLIB}) to Haskell and Isabelle
  (which are used by ATE systems).

@@

We are assuming that the ATE system acts in ``batch mode'' (accepting
definitions, producing conjectures, then halting), rather than running
continuously or interactively. This holds for all of the systems we have
analysed

@@

% We're assuming there is general agreement on what is "good maths", when
% actually this isn't at all clear-cut
